\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{booktabs}
\usepackage[round,authoryear]{natbib}
\usepackage{hyperref}

\title{Loss of SIA Loops to the Network-Dislocation Reservoir:\\
A Network-Only, Operator-Split Formulation for RadCluster\_2\_0}
\author{RadCluster\_2\_0 formulation note}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This note specifies the loop\,$\to$\,network loss channel that RadCluster\_2\_0
presently lacks, in a form that is (i)~physically correct for the
\emph{large, sessile} loops that actually feed the network, (ii)~non-redundant
with the SIA--SIA coalescence and Marian junction kernels already in the
model, and (iii)~implementable \emph{without} disturbing the constant-kernel
structure on which the C++ solver and its Woodbury preconditioner depend.
The key design commitment is operator splitting: the network density
$\rho_{\rm net}$ is held piecewise-constant within each integration segment
and refreshed between segments, so the Jacobian and preconditioner are never
re-derived.  The companion mass balances make the transfer exactly conservative
under the existing Frenkel-pair diagnostic.
\end{abstract}

\section{Introduction}

The absorption of irradiation-induced self-interstitial atom (SIA) loops
into the network dislocation structure is one of the central processes that
govern the long-time dose dependence of swelling, hardening, and
embrittlement in body-centred-cubic (BCC) metals and ferritic--martensitic
steels.  In the rate-theory tradition of Brailsford and
Bullough~\citep{Brailsford1972}, extended by Bullough, Eyre, and
Krishan~\citep{Bullough1975} and Mansur~\citep{Mansur1993}, the network is
folded into an effective sink strength $k_d^2$ that enters linearly in the
loss term $D_\alpha k_d^2 c$ of every diffusing-species balance.  The same
construction underlies essentially all production cluster-dynamics codes for
ferritic alloys and pure iron~\citep{GhoniemCho1979,HardouinDuparc2002,%
StollerGolubov2008,BarashevGolubov2009,KohnertWirth2018}: the network density
$\rho_{\rm net}$ enters as a \emph{static}, mean-field absorber whose biases
are set once at the start of the simulation.

RadCluster\_2\_0 inherits exactly this limitation.  The dislocation density
\texttt{rho\_d} is read once and enters only the fixed P4 sink
$\mathcal{D}_\alpha^d=Z_\alpha\,\rho_{\rm net}\,a^2\,\omega_\alpha^{\rm eff}$;
it never responds to the evolving loop population, and there is no channel
that removes loops \emph{from} the discrete population and adds them \emph{to}
the network.  Two consequences follow.  First, the model has no mechanism for
the experimentally robust \emph{saturation} of loop number density at high
dose, which physically arises when growing loops are swept into, or unfault
onto, the network.  Second, the largest loops---those past the SIA mobility
cutoff, and the entire sessile $\langle 100\rangle$ family---accumulate
without bound because they neither coalesce (no mobile partner) nor feel the
mobile-only fixed sink.

This note adds that missing channel.  Section~\ref{sec:design} states the
design constraints imposed by the existing solver.  Section~\ref{sec:physics}
gives the reframed, network-only loss term and its companion mass balances.
Section~\ref{sec:conservation} shows the transfer is exactly conservative under
the code's Frenkel-pair diagnostic.  Section~\ref{sec:impl} specifies the
operator-split implementation that avoids any Jacobian/preconditioner rework,
together with a phased plan and test gates.

\section{Design constraints from the existing solver}
\label{sec:design}

Three properties of the current code shape every modelling choice below.

\begin{enumerate}\itemsep2pt
  \item \textbf{Kernels are precomputed constants.}  Every reaction kernel is
        a \texttt{numpy} array built once and registered into the reaction
        admissibility graph (RAG); the C++ right-hand side treats sink
        coefficients as compile-time-constant per size.  A loss rate that
        depends on live aggregates ($N_\ell$, $\bar d_\ell$, $\rho_{\rm net}$)
        must therefore be recomputed at RHS-assembly time, and---crucially---%
        any feedback of $\rho_{\rm net}$ into the P4 sink diagonal must
        \emph{not} be made implicit, or the banded-plus-low-rank structure
        assumed by the Woodbury preconditioner is destroyed.
  \item \textbf{Two SIA loop characters, both with sessile populations.}  The
        glissile $\tfrac12\langle111\rangle$ family is mobile only up to
        \texttt{i\_mobile}; above it $D_n^{\rm eff}=0$.  The
        $\langle100\rangle$ family is sessile at all sizes.  The loss channel
        must remain \emph{nonzero for $D_n^{\rm eff}=0$}, because these are the
        loops that dominate network incorporation.
  \item \textbf{A live Frenkel-pair ledger.}  The diagnostic
        $\delta_{\rm FP}=|C_V-C_I-J^{\rm fix}_{I}+J^{\rm fix}_{V}|/(\cdots)$
        reads CVODE-integrated cumulative-sink state variables.  Any SIA
        content removed from the explicit loop inventory must be deposited in
        the cumulative fixed-sink integral $J^{\rm fix}_{I}$, or the diagnostic
        will report a spurious violation.
\end{enumerate}

\section{Network-only loss term}
\label{sec:physics}

\subsection*{Scope: why this is network-only}

The earlier draft of this note carried \emph{two} channels: loop--loop
($P_{\ell\ell}$) and loop--network ($P_{\ell d}$).  The loop--loop channel is
dropped here, deliberately.  RadCluster\_2\_0 already represents loop--loop
encounters through (a)~the same-character SIA--SIA coalescence kernel
$\mathcal{K}^{ii}$ and (b)~the Marian junction
$\tfrac12\langle111\rangle+\tfrac12\langle111\rangle\!\to\!\langle100\rangle$.
Adding a separate $P_{\ell\ell}$ loss would double-count those events.  This
note therefore models \emph{only} the transfer of loops to the pre-existing
network, $\Lambda_n^{\rm net}$, and leaves all loop--loop physics to the
coalescence/junction kernels.

\subsection*{Geometric length scales}

For a circular prismatic loop of $n$ SIAs of character $c\in\{111,100\}$ with
Burgers-vector magnitude $b_c$, the platelet area is $n\Omega/b_c$, so the
loop diameter is
\begin{equation}
  d_n^{(c)}=2\!\left(\frac{n\,\Omega}{\pi\,b_c}\right)^{1/2}.
  \label{eq:loop_diameter_from_n}
\end{equation}
The \emph{per-character} Burgers vector must be used; this is also the diameter
convention used by the TEM-diameter post-processing, and the two must agree.
The mean loop--network spacing is
\begin{equation}
  L_{\ell d}=\rho_{\rm net}^{-1/2},\qquad
  s_{\ell d}(n)=\max\!\left[\,b_c\,,\;L_{\ell d}-\tfrac12 d_n^{(c)}\right],
  \label{eq:loop_network_spacing}
\end{equation}
the edge-to-edge gap between a loop and the nearest network line, clamped from
below to prevent unphysical negative gaps.

\subsection*{Elastic interaction zone}

The elastic interaction range of a loop of size $n$ is
\begin{equation}
  R_{\rm int}(n)=\chi\,d_n^{(c)}
  \;+\;\underbrace{\xi\,b_c\sqrt{\dfrac{\mu\,\Omega\,n}{k_B T}}}_{\text{optional, small-$n$ floor}},
  \label{eq:elastic_interaction_zone}
\end{equation}
with $\chi=\mathcal{O}(1)$ the dimensionless geometric range parameter, $\mu$
the shear modulus, and $T$ temperature.  The optional second term keeps
$R_{\rm int}$ finite as $d_n\!\to\!0$; setting $\xi=0$ recovers
$R_{\rm int}=\chi d_n^{(c)}$.  \textbf{Recommended default:} $\xi=0$.  The
small-$n$ loops do not control saturation, and (see
\S\ref{sec:impl}) the extra parameter is not identifiable from
loop-density-versus-dose data.

\subsection*{Capture switch}

A smooth threshold gates the transfer on the geometric proximity of the loop
to the network,
\begin{equation}
  P_{\ell d}(n)
  =\tfrac12\!\left[1+\tanh\!\left(\frac{R_{\rm int}(n)-s_{\ell d}(n)}{\Delta_{\ell d}}\right)\right],
  \qquad \Delta_{\ell d}\sim b_c,
  \label{eq:P_loop_dislocation}
\end{equation}
so $P_{\ell d}\!\to\!0$ in the dilute limit ($L_{\ell d}\!\gg\!R_{\rm int}$)
and $P_{\ell d}\!\to\!1$ once the loop's elastic zone reaches the mean network
spacing.

\subsection*{Transfer frequency that survives sessile loops}

This is the central correction to the earlier draft.  The transfer rate is
\emph{not} tied to the loop's own diffusivity---if it were, it would vanish for
exactly the large/sessile loops that dominate incorporation (design constraint
2).  Loop incorporation proceeds by the network sweeping or climbing past a
stationary loop, so the relevant kinetics are set by the \emph{network}, not
the loop.  We write
\begin{equation}
  \Lambda_n^{\rm net}=\nu^{\rm net}\,P_{\ell d}(n),\qquad
  \nu^{\rm net}=v_{\rm net}\,\rho_{\rm net}\,w_c,
  \label{eq:Lambda_network}
\end{equation}
where $v_{\rm net}$ is the effective network climb/glide velocity, $\rho_{\rm
net}$ the network line density, and $w_c=\mathcal{O}(b_c)$ a capture width.
The product $v_{\rm net}\rho_{\rm net}$ is the area swept by the network per
unit volume per unit time; $w_c$ converts it to a capture probability per loop.
The climb velocity is itself irradiation-driven,
\begin{equation}
  v_{\rm net}=\frac{\Omega}{b\,}\,
  \bigl(Z_i\,\omega_i^{\rm eff}c_i-Z_v\,\omega_v^{\rm eff}c_v\bigr),
  \label{eq:vnet}
\end{equation}
i.e. proportional to the net biased point-defect flux already computed for the
P4 sink---so no new transport physics is introduced.  Because
Eq.~\eqref{eq:Lambda_network} contains \emph{no} factor of $D_n^{\rm eff}$, the
channel is active for sessile loops, as required.  A glissile-loop enhancement
may be added later as a separate, clearly-labelled term; it is not needed for
the saturation behaviour and is omitted here to keep the parameter set minimal.

\subsection*{Loss term}

The loop population loses members to the network at
\begin{equation}
  \left.\frac{dc_n}{dt}\right|_{\rm net}=-\,\Lambda_n^{\rm net}\,c_n,
  \label{eq:loop_network_loss}
\end{equation}
applied to both the $\tfrac12\langle111\rangle$ and $\langle100\rangle$
populations, for $n\ge n_{\rm inc}$ (an incorporation-onset size; loops below
$n_{\rm inc}$ are too small to register on the network and are excluded).

\subsection*{Companion mass balance for the network reservoir}

Each absorbed loop injects $n$ SIAs into the network and adds approximately one
loop circumference of dislocation line.  In atom-fraction variables (the code's
internal unit), number density is $c_n/\Omega$, so line-length and SIA-content
conservation require
\begin{align}
  \left.\frac{dN_{I}^{\rm net}}{dt}\right|_{\rm gain}
  &=\sum_{c}\sum_{n\ge n_{\rm inc}} n\,\Lambda_n^{\rm net}\,c_n^{(c)},
  \label{eq:N_sia_balance}\\[4pt]
  \frac{d\rho_{\rm net}}{dt}
  &=\frac{\pi}{\Omega}\sum_{c}\sum_{n\ge n_{\rm inc}} d_n^{(c)}\,\Lambda_n^{\rm net}\,c_n^{(c)}
  \;-\;\underbrace{K_{\rm rec}\,\rho_{\rm net}^{3/2}}_{\text{recovery}}.
  \label{eq:rho_net_balance}
\end{align}
The explicit $1/\Omega$ in Eq.~\eqref{eq:rho_net_balance} corrects a unit error
in the earlier draft: without it $\rho_{\rm net}$ is wrong by the factor
$\Omega\!\approx\!10^{-29}\,\mathrm{m^3}$.  The recovery sink
$K_{\rm rec}\rho_{\rm net}^{3/2}$ is \emph{mandatory}: with a gain term alone,
$\rho_{\rm net}$ grows without bound and the model merely relocates the
non-saturation problem from loops to the network.  $K_{\rm rec}$ sets the
steady-state network density and is the second of the two primary calibration
parameters.

\section{Conservation under the existing diagnostic}
\label{sec:conservation}

The Frenkel-pair diagnostic subtracts the cumulative SIA content lost to fixed
sinks, $J^{\rm fix}_{I}$.  The network \emph{is} a fixed sink, so the SIAs
carried away by absorbed loops, Eq.~\eqref{eq:N_sia_balance}, are routed into
the \emph{same} integral:
\begin{equation}
  \frac{dJ^{\rm fix}_{I}}{dt}\;\mathrel{+}=\;
  \sum_{c}\sum_{n\ge n_{\rm inc}} n\,\Lambda_n^{\rm net}\,c_n^{(c)}.
  \label{eq:ledger}
\end{equation}
With Eq.~\eqref{eq:ledger} in place, the swelling-identity diagnostic
$\delta_{\rm FP}$ is unchanged in form and remains below its $10^{-6}$ target:
every SIA leaving the explicit inventory $S_I=C_I$ reappears in $J^{\rm
fix}_{I}$, so $d(S_I+J^{\rm fix}_{I})/dt$ is untouched by the new channel.  No
modification to the $\delta_{\rm FP}$ formula is required---only the
accumulator feed.

\section{Implementation strategy}
\label{sec:impl}

\subsection*{The central commitment: operator-split, piecewise-constant
$\rho_{\rm net}$}

The hard part of this feature is \emph{not} the loss term---it is the feedback
of a now-dynamic $\rho_{\rm net}$ into the P4 sink.  Making $\rho_{\rm net}$ a
fully implicit state variable would make the P4 sink diagonal state-dependent
and break the banded-plus-low-rank Jacobian structure the Woodbury
preconditioner relies on, forcing a preconditioner/Jacobian rewrite.

\textbf{We avoid this entirely.}  $\rho_{\rm net}$ evolves on the dose
timescale---orders of magnitude slower than the point-defect relaxation that
sets solver stiffness---so it is held \emph{piecewise-constant within each
integration segment} and refreshed \emph{between} segments by operator
splitting:
\begin{enumerate}\itemsep2pt
  \item integrate the stiff cluster system over a segment with $\rho_{\rm net}$
        frozen, accumulating the gain integral
        $\int\!\frac{\pi}{\Omega}\sum d_n\Lambda_n c_n\,dt$ as an ordinary
        (non-stiff) quadrature;
  \item between segments, advance Eq.~\eqref{eq:rho_net_balance} explicitly
        (gain $-$ recovery) to update $\rho_{\rm net}$;
  \item call \texttt{rebuild\_rates()} so the P4 sink picks up the new
        $\rho_{\rm net}$ for the next segment.
\end{enumerate}
Within a segment all kernels remain constant; the Jacobian sparsity, the band
structure, and the Woodbury low-rank update are \emph{identical to today's}.
The adaptive driver \texttt{run\_adaptive} already segments the integration and
already calls \texttt{rebuild\_rates}, so the splitting reuses existing
machinery.  Only if a stiffness or accuracy problem is later demonstrated would
$\rho_{\rm net}$ be promoted to an implicit variable---and that decision is
deferred, not assumed.

The loss term itself, Eq.~\eqref{eq:loop_network_loss}, \emph{is} integrated
implicitly inside the segment.  Its coefficient $\Lambda_n^{\rm net}$ depends on
$\rho_{\rm net}$ (frozen within the segment) and on the live aggregates
$N_\ell,\bar d_\ell$ through $P_{\ell d}$.  This makes it a diagonal,
state-dependent loss on the SIA ladders---structurally identical to the
existing mobile P4 sink, and therefore \emph{inside} the band already; it adds
no new Jacobian coupling and no low-rank structure.

\subsection*{Combination with other sinks}

The network channel combines with the existing fixed sinks by the standard
additive-sink-strength rule (not the discontinuous $\min$ of length scales used
in the earlier draft): the total loss diagonal is the sum of the P4 fixed-sink
rate and $\Lambda_n^{\rm net}$.  This is smooth in the state and consistent with
how P4d/P4gb/P4p are already summed.

\subsection*{Phased plan and test gates}

\begin{description}
  \item[Phase 0 — physics freeze (this note).] Network-only; sweep-frequency
        $\nu^{\rm net}$ independent of $D_n^{\rm eff}$; additive sinks; recovery
        term; $1/\Omega$ units; per-character $b_c$.  Minimal parameter set
        $\{\chi,\;K_{\rm rec}\}$ with $\xi=0$, $n_{\rm inc}$ fixed.
  \item[Phase 1 — Python reference, $\rho_{\rm net}$ frozen.] Add
        $\Lambda_n^{\rm net}$ as a new SINK-class edge on \texttt{bulk-111} and
        \texttt{bulk-100}, kernel recomputed each RHS from current state in the
        GraphWalker; feed Eq.~\eqref{eq:ledger}.
        \emph{Gate:} $\chi\!\to\!0$ reproduces the current run; $\chi>0$ keeps
        $\delta_{\rm FP}<10^{-6}$.
  \item[Phase 2 — operator-split $\rho_{\rm net}$.] Implement the
        freeze/quadrature/\texttt{rebuild\_rates} cycle of
        \S\ref{sec:impl}.  \emph{Gate:} dislocation-line conservation holds to
        quadrature tolerance; loop number density now saturates with dose;
        Woodbury/Jacobian code unchanged.
  \item[Phase 3 — C++ mirror.] Replicate the RHS block in
        \texttt{rate\_kernels.cpp}; expose parameters in \texttt{parameters.h};
        wire \texttt{cpp\_bridge}.  \emph{Gate:} C++ and Python agree on the
        im5vm2-scale case to solver tolerance.
  \item[Phase 4 — bin-moment projection.] Project $\Lambda_n^{\rm net}c_n$ onto
        the bin moments using the active shape-function closure, mirroring the
        existing size-resolved sink projection.
  \item[Phase 5 — calibration.] Fit $\{\chi,K_{\rm rec}\}$ to the loop-density
        saturation in
        \texttt{docs/Database/FerriticSteels\_RadiationDatabase\_Combined.xlsx}.
\end{description}

\subsection*{Parameter identifiability}

A single loop-density-versus-dose saturation curve constrains roughly two
numbers: the \emph{onset dose} of saturation and the \emph{saturation level}.
These map naturally onto $\chi$ (when the geometric switch turns on) and
$K_{\rm rec}$ (the steady-state $\rho_{\rm net}$, hence the asymptotic sink
strength).  Introducing $\xi$, independent $\Delta_{\ell d}$, and a free
$w_c$ simultaneously would make the fit degenerate.  The model therefore ships
with $\xi=0$, $\Delta_{\ell d}=b_c$, $w_c=b_c$, $n_{\rm inc}$ fixed at the
mobility cutoff, and \emph{only} $\{\chi,K_{\rm rec}\}$ exposed for calibration;
the remaining knobs are promoted to free parameters only if a specific dataset
demands them.

\subsection*{New parameters}

The calibration parameters $\chi$ and $K_{\rm rec}$, and the fixed-but-tunable
$\xi,\Delta_{\ell d}/b_c,w_c/b_c,n_{\rm inc}$, are placed in the
\texttt{Model\_Parameters} sheet of \texttt{input\_parameters.xlsx} alongside
the existing $Z$-factors and capture radii, behind a master
\texttt{LOOP\_NETWORK\_LOSS} flag (default off) so the legacy behaviour is bit-for-bit
preserved when the channel is disabled.

\section{Limits and consistency}

\begin{itemize}\itemsep2pt
  \item \emph{Dilute limit} $\rho_{\rm net}\!\to\!0$: $L_{\ell d}\!\to\!\infty$,
        $P_{\ell d}\!\to\!0$, and both the loss term and the gain term vanish.
  \item \emph{Dense-network limit} $L_{\ell d}\!\lesssim\!R_{\rm int}$:
        $P_{\ell d}\!\to\!1$, $\Lambda_n^{\rm net}\!\to\!v_{\rm net}\rho_{\rm
        net}w_c$, the network-sweep-limited rate.
  \item \emph{Sessile loops} ($D_n^{\rm eff}=0$): channel remains active, by
        construction---this is the correction to the earlier draft.
  \item \emph{Channel disabled} (\texttt{LOOP\_NETWORK\_LOSS} off or $\chi=0$):
        $\Lambda_n^{\rm net}\equiv 0$, $\rho_{\rm net}$ static, and every
        equation reduces to the current model.
  \item \emph{Conservation}: SIAs removed from loops are accounted in
        $J^{\rm fix}_{I}$, Eq.~\eqref{eq:ledger}, so $\delta_{\rm FP}$ is
        preserved; dislocation line length is conserved through
        Eq.~\eqref{eq:rho_net_balance}.
\end{itemize}

\bibliographystyle{plainnat}
\bibliography{loop_network_loss}

\end{document}
